---
title: "COVID-19 USA Cases Analysis"
author: "Guy Singer"
date: "4/26/2022"
output:
  pdf_document: default
  html_document: default
urlcolor: blue
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries Used
```{r setup2, include=TRUE}
library(tidyverse)
library(lubridate)
```

## Introduction
This project examines COVID-19 case data to model the relationship between date of the year and new cases, with the hope of robustly predicting new cases based on the time of year.

## Importing the Data
The time series data in this project are imported from the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, and can be found here: <https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series>

```{r get_data, message = FALSE}

## get current Data
US_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
US_deaths <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

```

## Tidying the Data
I start by cleaning this data by removing columns we will not be using, such as Lat and Long, and changing variable names to be more friendly for R, such as 'date'.

```{r tidy_data2, message = FALSE}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)

US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases))

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases))%>%
  mutate(case_accel = new_cases / lag(new_cases))

US_totals
summary(US_totals)
```

## Two Visualizations of the Data
I then examine and analyze the data using two different visualization techniques. These visualizations will assist me in proposing a model appropriate to the data. The first visualization is an bar plot of the acceleration of new cases across the united states as a function of time.

```{r tidy_visualization1, warning = FALSE, message = FALSE, fig.width=6, fig.height=2.5}

US_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y=case_accel)) +
  geom_bar(stat='identity', fill = "steelblue") +
  scale_y_log10()+
  labs(title = "Acceleration of New COVID-19 Cases in the USA (Log Scale)", y = NULL)

```

The second visualization is a simple area plot charting the new case counts in California as a function of time.

```{r tidy_visualization2, warning = FALSE, message = FALSE, fig.width=6, fig.height=2.5}

US_by_state %>%
  filter (Province_State == "California")%>%
  filter(cases>0) %>%
  ggplot(aes(x = date, y=new_cases)) +
  geom_area(stat='identity', fill = "steelblue") +
  scale_y_log10()+
  labs(title = "New COVID-19 Cases in California (Log Scale)", y = NULL)
```


## Modelling the Data
In order to provide a good model, I want to account for non-linear relationships. Therefore, I build a quadratic regression model that is effectively a linear model in two variables, with the second variable being a square of the first.
```{r data_model, warning = FALSE, message = FALSE, fig.width=6, fig.height=2.5}

US_state_totals <- US_by_state %>%
  group_by(Province_State)%>%
  summarize(deaths = max(deaths), cases =max(cases),
            population = max(Population),
            cases_per_thou = 1000* cases/population,
            deaths_per_thou = 1000* deaths/population,
            cases_per_thou_2 = cases_per_thou^2 ) %>%
  filter(cases>0, population >0)


quad.mod <- lm(deaths_per_thou ~ cases_per_thou + cases_per_thou_2, data = US_state_totals)
summary (quad.mod)

US_tot_w_pred <- US_state_totals %>% mutate (pred=predict(quad.mod))

US_tot_w_pred %>%
  ggplot()+
  geom_point(aes(x = cases_per_thou, y=deaths_per_thou), color = "blue")+
  geom_point(aes(x=cases_per_thou, y = pred), color = "red")
```

These results, demonstrating that although my faux-quadratic model does a reasonable job at predicting the data, there are still many data points that do not line up with the model. Therefore, there must be additional variable not being considered in the model. The question of what these variables may be and how they are computed in the model merits further examination.

## Conclusions
The conclusions of this experiment suggest that cases per thousand are highly correlated with deaths per thousand, and that this relationship can be fairly robustly modeled using quadratic regression.
Potential sources of bias in this experiment are many, and should be mentioned. The collection of COVID-19 case data has been shown to suffer from a number of biases, including those related to differing data collection rates in differing socioeconomic communities. Access to testing and diagnosis tools are not evenly distributed and likely introduce significant bias to the model. Additionally, as the author I bring personal biases, such as those relating to socioeconomic class, that I have attempted to assuage by implementing a controlled experiment.